---
title: "Prioritizing candidates for insects"
author: "Fabrício Almeida-Silva"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,
                      warning=FALSE, 
                      message=FALSE,
                      tidy.opts=list(width.cutoff=60),
                      tidy=TRUE)
```

## Candidate gene mining

Now, we will mine candidate genes with {cageminer}. To start, let's get the guide genes.

```{r load_data}
library(ggplot2)
library(here)
library(cageminer)
library(tidyverse)

# Load SNP ranges
load(here("products", "result_files", "snp_granges.rda"))

# Create a gene-based GRanges object
gff <- rtracklayer::import(here("data", "PLAZA_selected.transcripts.gff.gz"))
gff_gene <- gff[gff$type == "gene", ]
rm(gff)
```

```{r get_guides}
#----Get biotic stress-related genes from Almeida-Silva e Venancio 2021----
guides <- readr::read_tsv("https://raw.githubusercontent.com/almeidasilvaf/SoyFungi_GWAS_GCN/main/products/tables/sup_table3.tsv")
write_tsv(guides, 
          here("products", "tables", "sup_table3.tsv"))

# EDA: Number of genes per class
guides %>%
  janitor::tabyl(Description) %>%
  arrange(-n) %>%
  head(n=15)
```

Find high-confidence candidates for insects.

```{r insects}
library(BioNERO)
library(SummarizedExperiment)
set.seed(1)

#----Load and filter expression data----
load("~/Dropbox/Atlas/atlasv2_tpm.rda")
exp_insect <- atlas_tpm[, atlas_tpm$Stress_info == "insect" &
                   !is.na(atlas_tpm$Stress_info)]
rm(atlas_tpm)

#----Infer GCNs for insects----
exp_insect <- exp_preprocess(exp_insect, min_exp=5, Zk_filtering=FALSE)
metadata_insect <- colData(exp_insect) %>%
  as.data.frame() %>%
  mutate(Pathogen = str_sub(Pathogen, 1, 3)) %>%
  mutate(annot = paste(Pathogen, Sample_description, sep = "_")) %>%
  select(annot)
colData(exp_insect) <- DataFrame(metadata_insect)

sft_insect <- SFT_fit(exp_insect, net_type="unsigned", cor_method="pearson")
gcn_insect <- exp2gcn(exp_insect, net_type="unsigned", cor_method="pearson",
                     SFTpower=sft_insect$power)

# Cleaning up
source(here::here("code", "utils.R"))
workspace_size()
gcn_insect$adjacency_matrix <- NULL
gcn_insect$dendro_plot_objects <- NULL
gcn_insect$moduleColors <- NULL
gcn_insect$correlation_matrix <- round(gcn_insect$correlation_matrix, 2)

#----Mine candidates----
# Aphis glycines
conditions_agl <- c(
  "Agl_stress_res_48h", "Agl_stress_res_6h", "Agl_stress_res_12h",
  "Agl_stress_48HAI_susceptible", "Agl_stress_96HAI_susceptible",
  "Agl_stress_48HAI_antibiotic", "Agl_stress_96HAI_antibiotic",
  "Agl_stress_48HAI_antixenotic", "Agl_stress_96HAI_antixenotic",
  "SBA_sus_24h", "SBA_res_24h",
  "SBA_sus_48h", "SBA_res_48h"
)
mining_aglycines <- mine_candidates(
  gene_ranges = gff_gene, 
  marker_ranges = snp_grangeslist$Aglycines,
  exp = exp_insect,
  gcn = gcn_insect,
  guides = guides,
  metadata = metadata_insect,
  sample_group = conditions_agl
)
length(unique(mining_aglycines$gene))

# Spodoptera litura
conditions_sli <- c(
  "Sli_stress_NNcultivar", "Sli_stress_WXcultivar"
)
exp_sli <- exp_insect[, grepl("Sli.*", exp_insect$annot)]
mining_slitura <- mine_candidates(
  gene_ranges = gff_gene, 
  marker_ranges = snp_grangeslist$Slitura,
  exp = exp_sli,
  gcn = gcn_insect,
  guides = guides,
  metadata = metadata_insect,
  sample_group = conditions_sli
)
length(unique(mining_slitura$gene))

# Save all candidates
save(mining_aglycines, 
     file = here("products", "result_files", "aglycines_minedcand.rda"),
     compress="xz")

save(mining_slitura, 
     file = here("products", "result_files", "slitura_minedcand.rda"),
     compress="xz")
```

Now, let's score genes using cageminer's scoring function.

```{r scoring}
cand_unique <- list(
  Aglycines = unique(mining_aglycines$gene),
  Slitura = unique(mining_slitura$gene)
)

#----Get candidates that are hubs----
hubs <- get_hubs_gcn(exp_insect, gcn_insect)
candidates_hubs <- lapply(cand_unique, function(x) {
  return(hubs[hubs$Gene %in% x, ])
})

#----Get candidates that are TFs----
tfs <- as.data.frame(readr::read_tsv("http://planttfdb.gao-lab.org/download/TF_list/Gma_TF_list.txt.gz"))[, c(2,3)]
candidates_tfs <- lapply(cand_unique, function(x) {
  y <- tfs[tfs$Gene_ID %in% x, ]
  return(y)
})

#----Score genes----
hubs <- hubs$Gene
tfs <- tfs$Gene_ID

aglycines_score <- score_genes(mining_aglycines, hubs, tfs)
slitura_score <- score_genes(mining_slitura, hubs, tfs)

#----Export best candidates as a table----
dfexport_candidates <- rbind(
  aglycines_score %>% mutate(species = "A. glycines") %>% select(1,5,6),
  slitura_score %>% mutate(species = "S. litura") %>% select(1,5,6)
)

write_tsv(dfexport_candidates, 
          file = here("products", "tables", "top_candidates_insects.tsv"))
```

Exporting descriptions for mined candidates.

```{r description}
gmax_des <- readr::read_tsv("https://ftp.psb.ugent.be/pub/plaza/plaza_public_dicots_05/Descriptions/gene_description.gma.csv.gz", skip=8)
names(gmax_des) <- c("Gene", "description_type", "description")

cand_description <- Reduce(rbind, list(
  aglycines = gmax_des[gmax_des$Gene %in% cand_unique$Aglycines, ] %>%
    mutate(species = "Aglycines"),
  slitura = gmax_des[gmax_des$Gene %in% cand_unique$Slitura, ] %>%
    mutate(species = "Situra")
))
write_tsv(cand_description, 
          file = here("products", "tables", "mined_candidates_description.tsv"))
```


## Export module enrichment and network file for shiny app development

```{r export_files}
options(timeout = 600)

# Get GO annotation
go_annotation <- as.data.frame(
  readr::read_tsv("https://ftp.psb.ugent.be/pub/plaza/plaza_public_dicots_05/GO/go.gma.csv.gz", skip=8)
)
go_annotation <- go_annotation[, c(1,8)]
names(go_annotation) <- c("Gene", "GO_description")

# Get Interpro annotation
interpro_annotation <- as.data.frame(
  readr::read_tsv("https://ftp.psb.ugent.be/pub/plaza/plaza_public_dicots_05/InterPro/interpro.gma.csv.gz", skip=8)
)
interpro_annotation <- interpro_annotation[, c(1,4)]

# Get MapMan annotation
mapman <- readr::read_tsv("https://ftp.psb.ugent.be/pub/plaza/plaza_public_dicots_05/MapMan/mapman.gma.csv.gz", skip=8)
mapman_annotation <- as.data.frame(mapman[, c(3,5)])
names(mapman_annotation) <- c("Gene", "Description")

enrich_mapman <- module_enrichment(
       net = gcn_insect,
       background_genes = rownames(exp_insect),
       mapman_annotation
)

enrich_go <- module_enrichment(
       net = gcn_insect,
       background_genes = rownames(exp_insect),
       go_annotation
)

enrich_interpro <- module_enrichment(
       net = gcn_insect,
       background_genes = rownames(exp_insect),
       interpro_annotation
)

enrichment_results <- Reduce(rbind, list(
  mapman = enrich_mapman[, c(1,5,7)] %>% mutate(Category = "MapMan"),
  go = enrich_go[, c(1,5,7)] %>% mutate(Category = "GO"),
  interpro = enrich_interpro[, c(1,5,7)] %>% mutate(Category = "Interpro")
))


genes_modules <- gcn_insect$genes_and_modules
genes_modules <- genes_modules[genes_modules$Modules != "grey", ]
degree <- gcn_insect$kIN

edges <- lapply(unique(genes_modules$Modules), function(x) {
  y <- get_edge_list(gcn_insect, module = x, filter=TRUE,
                     method="min_cor", rcutoff = 0.4)
  return(y)
})
names(edges) <- unique(genes_modules$Modules)

greygenes <- genes_modules[genes_modules$Modules == "grey", 1]
cor_matrix <- gcn_insect$correlation_matrix[
  !(rownames(gcn_insect$correlation_matrix) %in% greygenes),
  !(colnames(gcn_insect$correlation_matrix) %in% greygenes)
]
cor_matrix[cor_matrix < 0.2 & cor_matrix > -0.2] <- NA
edgelist <- cormat_to_edgelist(cor_matrix)

# Export files
save(enrichment_results, 
     file = here("products", "result_files", "shiny_insect_enrichment.rda"),
     compress = "xz")

save(genes_modules, 
     file = here("products", "result_files", "shiny_insect_gmodules.rda"), 
     compress="xz")

save(degree, 
     file = here("products", "result_files", "shiny_insect_degree.rda"), 
     compress="xz")

save(edges, 
     file = here("products", "result_files", "shiny_insect_edgelists.rda"), 
     compress="xz")

save(hubs, 
     file = here("products", "result_files", "shiny_insect_hubs.rda"), 
     compress="xz")
```

## Session information

This document was created under the following conditions:

```{r session_info} 
sessionInfo()
```



