---
title: "Summary stats of prioritized candidates"
author: "Fabrício Almeida-Silva"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE, 
                      message = FALSE,
                      tidy.opts = list(width.cutoff = 60),
                      tidy = TRUE)
```

## Functional classification of candidates

The file "all_mined_candidates" was manually modified to include descriptions for all genes and their role in plant defense. Let's plot data now.

```{r candidate_classification_stats}
library(tidyverse)
library(here)
cpal <- c("darkgoldenrod3", "dimgrey", "darkgreen")

all_cand_des <- read.csv(
    here("products", "tables", "all_mined_candidates_and_description.txt"),
    header = TRUE, sep = "\t"
)

freq_abs_cand <- all_cand_des %>%
    janitor::tabyl(Role, Species) %>%
    mutate(Role = stringr::str_replace_all(Role, "SM metabolism", 
                                           "Specialized metabolism")) %>%
    reshape2::melt()

freq_cand <- freq_abs_cand %>%
    group_by(variable) %>%
    mutate(percentage = value / sum(value))

p_cand <- ggplot(freq_cand) +
    geom_bar(aes(x=percentage, y=Role, fill = variable), stat="identity",
             show.legend=FALSE) +
    ggplot2::scale_fill_manual(values = cpal) +
    facet_wrap(~variable, ncol=5) +
    theme_bw() +
    labs(x="Relative frequency", y="")

ggsave(filename = here("products", "plots", "candidates_classification.png"),
       p_cand, width = 9, height=3)

# Combine figures
library(magick)
library(patchwork)
f1 <- magick::image_read(here("products", "plots", "biorender_plant_defense_to_pests.png"))
f2 <- magick::image_read(here("products", "plots", "candidates_classification.png"))

final_fig <- image_append(c(image_scale(f1, "2700x2020!"), f2), 
                          stack=TRUE)
magick::image_write(
    final_fig, 
    path = here("products", "plots", "candidates_classification_final.png")
)
```

Show candidates in a Venn Diagram.

```{r venn_plot}
# Load data
load(here("products", "result_files", "aglycines_minedcand.rda"))
load(here("products", "result_files", "hglycines_minedcand.rda"))
load(here("products", "result_files", "slitura_minedcand.rda"))


venn_pdata <- list(
  Aglycines = unique(mining_aglycines$gene),
  Slitura = unique(mining_slitura$gene),
  Hglycines = unique(mining_hglycines$gene)
)

# Plot venn diagram using the VennDiagram package
library(VennDiagram)
venn.diagram(
        x = venn_pdata,
        category.names = c("A. glycines", "S. litura", "H. glycines"),
        filename = here("products", "plots", "venn_diagram_candidates.png"),
        output=TRUE,
        
        # Output features
        imagetype="png" ,
        height = 1600, 
        width = 1600, 
        resolution = 600,
        
        # Circles
        lwd = 2,
        lty = 'blank',
        fill = c("darkgoldenrod3", "darkgreen", "dimgrey"),
        
        # Numbers
        cex = 0.5,
        fontface = "bold",
        fontfamily = "sans",
        
        # Set names
        cat.cex = 0.5,
        cat.fontface = "italic",
        cat.default.pos = "outer",
        cat.fontfamily = "sans",
        margin = 0.2
)
```

## Comparing to fungi resistance-related genes

```{r}
cgregata <- paste0(tempdir(), "/cgregata.rda")
download.file("https://github.com/almeidasilvaf/SoyFungi_GWAS_GCN/blob/main/products/result_files/cgregata_minedcand.rda?raw=true", destfile = cgregata)

fgraminearum <- paste0(tempdir(), "/fgraminearum.rda")
download.file("https://github.com/almeidasilvaf/SoyFungi_GWAS_GCN/blob/main/products/result_files/fgraminearum_minedcand.rda?raw=true", destfile = fgraminearum)

fvirguliforme <- paste0(tempdir(), "/fvirguliforme.rda")
download.file("https://github.com/almeidasilvaf/SoyFungi_GWAS_GCN/blob/main/products/result_files/fvirguliforme_minedcand.rda?raw=true", destfile = fvirguliforme)

mphaseolina <- paste0(tempdir(), "/mphaseolina.rda")
download.file("https://github.com/almeidasilvaf/SoyFungi_GWAS_GCN/blob/main/products/result_files/mphaseolina_minedcand.rda?raw=true", destfile = mphaseolina)

ppachyrhizi <- paste0(tempdir(), "/ppachyrhizi.rda")
download.file("https://github.com/almeidasilvaf/SoyFungi_GWAS_GCN/blob/main/products/result_files/ppachyrhizi_minedcand.rda?raw=true", destfile = ppachyrhizi)


load(cgregata)
load(fgraminearum)
load(fvirguliforme)
load(mphaseolina)
load(ppachyrhizi)

venn_pdata_comp <- list(
  Aglycines = unique(mining_aglycines$gene),
  Slitura = unique(mining_slitura$gene),
  Hglycines = unique(mining_hglycines$gene),
  Cgregata = unique(mining_cgregata$gene),
  Fgraminearum = unique(mining_fgraminearum$gene),
  Fvirguliforme = unique(mining_fvirguliforme$gene),
  Mphaseolina = unique(mining_mphaseolina$gene),
  Ppachyrhizi = unique(mining_ppachyrhizi$gene)
)

# Plot venn diagram using the VennDiagram package
library(ComplexHeatmap)
upset_matrix <- make_comb_mat(list_to_matrix(venn_pdata_comp))

ht <- draw(
  UpSet(
    upset_matrix,
    set_order = c("Hglycines", "Aglycines", "Slitura",
                  "Fvirguliforme", "Fgraminearum", "Cgregata",
                  "Mphaseolina", "Ppachyrhizi")
  )
)
od <- column_order(ht)
cs <- comb_size(upset_matrix)
decorate_annotation("intersection_size", {
    grid.text(cs[od], x = seq_along(cs), y = unit(cs[od], "native") + unit(2, "pt"), 
        default.units = "native", just = "bottom", gp = gpar(fontsize = 8))
})
# Saved as PDF, 6 x 5 inches
```

## Comparing to previous studies

To conclude, we will compare the high-confidence candidate genes we found
with potential resistance genes identified in previous studies.

First, let's manually identify each gene and their respective
studies and look for matches between reported genes and our list of prioritized
candidates.

```{r known_resistance_genes}
# H. glycines
## Important loci
rhg1 <- c(
    "Glyma.18G022400", "Glyma.18G022500", "Glyma.18G022600", "Glyma.18G022700"
)

rhg4 <- "Glyma.08G107700"
sum(c(rhg1, rhg4) %in% all_cand_des$Gene)

## Loci reported in previous GWAS papers
liu_2019 <- c(
    "Glyma.09G054000", "Glyma.16G065700", "Glyma.18G192200"
)

zhang_2017 <- c(
    "Glyma.08G108900", "Glyma.08G303900", "Glyma.11G151500",
    "Glyma.13G175200", "Glyma.17G010200", "Glyma.18G022700",
    "Glyma.18G95000"
)

chang_2016 <- c(
    "Glyma.18g023500", "Glyma.20g193500"
)

vuong_2015 <- c(
    NA
)

zhao_2017 <- c(
    "Glyma.07G195400", "Glyma.07G196000", "Glyma.08G100100", 
    "Glyma.18G193200", "Glyma.18G193300", "Glyma.18G193400",
    "Glyma.07G190900", "Glyma.07G191000", "Glyma.07G191100",
    "Glyma.07G191200", "Glyma.07G191500", "Glyma.07G195500",
    "Glyma.08G099400", "Glyma.08G099700", "Glyma.08G100700",
    "Glyma.08G100800", "Glyma.08G200800", "Glyma.08G200800",
    "Glyma.08G201000", "Glyma.08G201100", "Glyma.14G048600",
    "Glyma.14G049500"
)

tran_2019 <- c(
    "Glyma.07g194200", "Glyma.07g194400", "Glyma.07g194800",
    "Glyma.07g195000", "Glyma.07g195100", "Glyma.07g195300",
    "Glyma.07g195400", "Glyma.07g195500", "Glyma.07g195700",
    "Glyma.07g195900", "Glyma.07g196000", "Glyma.07g196500",
    "Glyma.07g199000", "Glyma.07g200100", "Glyma.07g199500",
    "Glyma.07g199700", "Glyma.07g199900", "Glyma.10g172700",
    "Glyma.10g172800", "Glyma.10g172900", "Glyma.10g173000",
    "Glyma.10g173100", "Glyma.10g173300", "Glyma.10g173400"
)

bao_2014 <- c(
    "Glyma.18g227700"
)

sum(all_cand_des$Gene %in% c(
    bao_2014, tran_2019, zhao_2017, vuong_2015, chang_2016, zhang_2017, liu_2019
))

# S. litura
liu_2016 <- c(
    "Glyma.07g051500", "Glyma.07g055200", "Glyma.06g189600", "Glyma.06g190800"
)

sum(all_cand_des$Gene %in% c(
    liu_2016
))

# A. glycines
natukunda_2019 <- c(
    "Glyma.12G075100", "Glyma.10G015900", "Glyma.10G297600"
)

chang_hartman_2017 <- c(
    "Glyma.07G117400", "Glyma.07G108200", "Glyma.07G108400",
    "Glyma.07G110700"
)

hanson_2018 <- c(
    NA
)

sum(all_cand_des$Gene %in% c(
    natukunda_2019, chang_hartman_2017, hanson_2018
))

all_cand_des |>
    filter(Gene %in% c(natukunda_2019, chang_hartman_2017, hanson_2018))
```

Only 1 prioritized gene for resistance to *A. glycines* has been previously
reported (Glyma.07G117400, Hartman et al., 2017). This gene encodes
a leucine-rich repeat receptor kinase (LRR-RK), a receptor that is typically
involved in the recognition of non-self molecular patterns.

For *H. glycines*, we identified one gene (Glyma.18G022400) from the Rhg1 locus,
which is known to confer resistance to cyst nematodes not only in soybean,
but also in other plant species (Cook et al., 2012; Butler et al., 2019).

Finally, we will look for prioritized candidate genes that are in known
QTL regions identified by biparental crossing and QTL mapping. For that, we will
extract known QTL for each trait from SoyBase.

```{r}
library(rvest)

# Define function to obtain QTL IDs for a trait
get_QTL_list_from_soybase <- function(qtl_url) {
    
    qtl_df <- read_html(
        qtl_url
    ) |>
        html_table() |>
        pluck(2)
    
    return(qtl_df)
}

# Define function to extract QTL coordinates from QTL IDs
get_QTL_coordinates <- function(QTL_id) {
    
    # Read HTML page and extract text with positions of nearest markers
    url <- paste0(
        "https://www.soybase.org/SeqMapSearch/GbrowseSearch/SeqGenSearch4.php?qtlname=",
        gsub(" ", "%20", QTL_id)
    )
    
    page <- read_html(url) |>
        html_elements(".sb_main") |>
        html_elements(xpath = "/html/body/div[3]/div/text()") |>
        html_text2() 
    
    # Extract text with pattern "GmX:NNN"
    fpage <- page[grepl("position", page)]
    positions <- gsub(".*position: ", "", fpage)
    
    # Get chromosome number, start, and end positions
    chr <- unique(gsub(":.*", "", positions))
    
    coords <- NULL
    if(length(chr) > 0) {
        position_number <- sort(as.numeric(gsub(".*:", "", positions)))
        start <- position_number[1]
        end <- position_number[2]
        
        coords <- data.frame(seqnames = chr, start = start, end = end)
    } 
    
    return(coords)
}

# Get QTL coordinates for each trait
## H. glycines (SCN)
qtl_ids_scn <- get_QTL_list_from_soybase(
    "https://www.soybase.org/search/index.php?searchterm=SCN&list=bi_parental_qtl_listview"
)
qtl_ids_scn <- qtl_ids_scn[, 3, drop = TRUE]
qtl_ids_scn <- qtl_ids_scn[startsWith(qtl_ids_scn, "SCN ")]

qtl_coord_scn <- Reduce(rbind, lapply(qtl_ids_scn, get_QTL_coordinates)) |>
    mutate(seqnames = str_replace_all(seqnames, "Gm", "Chr"))

## S. litura (common cutworm)
qtl_ids_sli <- paste("Common cutworm", c(
    "1-1", "1-2", "2-1", "2-2"
))
qtl_coord_sli <- Reduce(rbind, lapply(qtl_ids_sli, get_QTL_coordinates)) |>
    mutate(seqnames = str_replace_all(seqnames, "Gm", "Chr"))

## A. glycines (soybean aphid)
qtl_ids_agly <- paste("Rag", c(
    "2-1", "2-2", "3-1", "3-2", "4-1", "4-2", "4-3"
))
qtl_coord_agly <- Reduce(rbind, lapply(qtl_ids_agly, get_QTL_coordinates)) |>
    mutate(seqnames = str_replace_all(seqnames, "Gm", "Chr"))

# Look for prioritized candidates in QTL regions
library(plyranges)

gene_ranges <- rtracklayer::import(
    here("data", "PLAZA_selected.transcripts.gff.gz")
) |>
    filter(type == "gene") |>
    select(gene_id)

## H. glycines
prioritized_genes_in_qtls_scn <- gene_ranges |>
    filter(gene_id %in% all_cand_des[all_cand_des$Species == "Hglycines", "Gene"]) |>
    join_overlap_intersect(as_granges(qtl_coord_scn)) |>
    as.data.frame() |>
    dplyr::pull(gene_id) |>
    unique()

length(prioritized_genes_in_qtls_scn)

## S. litura
prioritized_genes_in_qtls_sli <- gene_ranges |>
    filter(gene_id %in% all_cand_des[all_cand_des$Species == "Slitura", "Gene"]) |>
    join_overlap_intersect(as_granges(qtl_coord_sli)) |>
    as.data.frame() |>
    dplyr::pull(gene_id) |>
    unique()

length(prioritized_genes_in_qtls_sli)

## A. glycines
prioritized_genes_in_qtls_agly <- gene_ranges |>
    filter(gene_id %in% all_cand_des[all_cand_des$Species == "Aglycines", "Gene"]) |>
    join_overlap_intersect(as_granges(qtl_coord_agly)) |>
    as.data.frame() |>
    dplyr::pull(gene_id) |>
    unique()

length(prioritized_genes_in_qtls_agly)
```

The results per species are as follows:

- *H. glycines* (SCN): 41% (80/194) of prioritized genes in known QTLs.
- *S. litura*: 0% (0/7) of the prioritized genes in known QTLs.
- *A. glycines* (soybean aphid): 4% (6/171) of the prioritized genes in known QTLs.

QTLs were obtained from biparental mapping.

## Session information

This document was created under the following conditions:

```{r session_info} 
sessionInfo()
```



